A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2024-09-18 manuscript.qmd
testing overleaf setup
first changes

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
abstract: |
  Purpose: According to current guidelines for elective nodal irradiation of oropharyngeal squamous cell carcinoma (OPSCC) patients, large parts of the contralateral lymphatic system are included in the elective clinical target volume (CTV-N), even for lateralized tumors without clinical lymph node involvement in the contralateral neck. In this work, we present a probabilistic model for bilateral lymphatic tumor progression in OPSCC to predict the personal risk for occult disease in any lymph node level (LNL), given the patient's clinical lymph node involvement, T-stage, and lateralization of the primary tumor.

  Methods: We extend a hidden markov model for lymphatic tumor progression, which was previously developed for ipsilateral lymph node involvement, to the contralateral neck. The model represents each of the LNLs I, II, III, IV, V, and VII of both sides of the neck as a hidden binary random variable with states healthy and involved. LNLs are connected to the tumor and among each other via arcs that correspond to spread probabilities. These spread probability rates are learned via Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC patients.

  Results: The model is able to describe the data on lymph node involvement well with a small number of interpretable parameters. Midline extension of the primary tumor is the main risk factor for contralateral involvement. In addition, the risk of occult metastases in contralateral lymph node level increases with more advanced T-stage and more severe ipsilateral involvement. The probability of involvement in contralateral level III is very low if the upstream level II is clinically negative. Similarly, the probability of involvement in contralateral level IV is very low if the upstream level III is clinically negative.

  Conclusions: The model may guide personalized volume reduction of the elective CTV-N. It suggests that the contralateral neck may be spared from elective irradiation for lateralized tumors not crossing the midline. In patients with primary tumors crossing the midline but clinically negative contralateral neck, the contralateral elective irradiation may be limited to level II.
---

```{python}
#| echo: false
import re

import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from matplotlib import ticker
from matplotlib.colors import LinearSegmentedColormap
import upsetplot

from lymph import models
from lyscripts import utils
from lyscripts.plot.utils import COLORS

from scripts import shared, paths
from scripts.shared import COL, get_lnl_cols

simple_model = shared.get_model("simple", load_samples=True)
```

```{python}
#| tags: [parameters]
#| echo: false

# width of figures. May depend on number of text columns
full = 17 # cm
half = full
# half =  8 # cm
```

# Introduction

When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted. Per current guidelines, large volumes of the neck are part of the the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009]. This is to minimize the risk of regional recurrences due to untreated microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET). However, minimizing the risk of missing occult disease in the lymph drainage region must be balanced against the toxicity related to unnecessary treatment of healthy tissue.

The aforementioned guidelines for the definition of the CTV-N are based on anatomically defined lymph node levels (LNL) [@gregoire_delineation_2014] and the overall prevalence of lymph node metastases in the LNLs. Based on this, they recommend extensive irradiation of both sides of the neck for a majority of patients. However, this prevalence the guidelines are based on does not take into account an individual patient's state of tumor progression, which may have a great impact on a patient's personal risk for occult disease in a given LNL. For example, a patient with substantial nodal involvement in the ipsilateral side and an advanced primary tumor that crosses the mid-sagittal plane would receive extensive elective irradiation of the contralateral LNLs II, III, and IVa. However, if the patient presented with no clinically detectable nodal disease (cN0) and only a small, clearly lateralized T1 tumor, the same contralateral CTV-N would be recommended [@biau_selection_2019].

To better quantify this personalized risk of occult disease we have previously developed an intuitive, probabilistic hidden Markov model [@ludwig_hidden_2021;@ludwig_modelling_2023] which was originally based on a conceptually similar Bayesian network model [@pouymayou_bayesian_2019]. So far, these models were limited to the ipsilateral side of the neck. The main contribution of this work will be to extend the formalism to include risk predictions for the contralateral side as well. We believe this holds the potential to reduce the electively irradiated volumes in the contralateral neck for patients whose individual risk for occult disease is low. In turn, this may substantially reduce the toxicity from irradiation and the associated decrease in quality of life.

The main contributions of this paper are as follows:

1. In section @sec-data we present a multi-centric dataset on lymph node involvement in {{< var data.num_patients >}} OPSCC patients. Based on the dataset the main risk factors for contralateral lymph node involvement are identified and the requirements for the bilateral model extension are described (@sec-requirements).

2. In @sec-ext-to-contra we present a bilateral HMM of lymphatic progression that accounts for the T-stage lateralization of the primary tumor as well as clinical involvement as risk factors for occult contralateral involvement. In @sec-methods we describe how the model training and the computational experiments were set up.

3. In the results in @sec-results, we demonstrate that the model is able to describe the patterns of contralateral lymph node involvement observed in the dataset, and we apply the model to estimating risk of occult disease for typical patients. In @sec-discussion, implications for volume-deescalated radiotherapy are discussed.

@@

To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023]. It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status. Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview. The data from the Inselspital Bern and the Centre Léon Bérard only consist of patients who received a neck dissection. Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset. The majority of patients from the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated with definitive radiotherapy. For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.

```{python}
#| echo: false
#| label: tbl-data-overview
#| tbl-cap: Overview over the five datasets from four different institutions used to train and evaluate our model. Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received some form of neck dissection, the N0 portion of patients, what percentage presented with early T-category, and the prevalence of primary tumor midline extension. For a much more detailed look at the data, visit [lyprox.org](https://lyprox.org).
from typing import Literal

def align(
  column: pd.Series,
  where: Literal["right", "left", "center"],
) -> list[str]:
  """Align a column."""
  return [f"text-align: {where}"] * len(column)

def right_align(column: pd.Series) -> list[str]:
  """Right align column."""
  return align(column, where="right")

def highlight(column: pd.Series, mapping: dict) -> list[str]:
  """Color based on `mapping`."""
  colors = column.map(mapping)
  return colors.map(lambda x: f"color: {x}")

def highlight_bool(
  column: pd.Series,
  c_false: str = COLORS["green"],
  c_true: str = COLORS["red"],
) -> list[str]:
  """Make cell read (`False`) or green (`True`)."""
  return highlight(column, mapping={True: c_true, False: c_false})

def highlight_t_stage(
  column: pd.Series,
  c_early: str = COLORS["green"],
  c_late: str = COLORS["red"],
) -> list[str]:
  """Highlight `early` and `late`."""
  is_early = column == "early"
  return highlight_bool(is_early, c_false=c_late, c_true=c_early)

col_map = {
  COL.inst: "Institution",
  COL.age: "Age",
  COL.nd: "Neck Dissection",
  COL.t_stage: "T-category",
  COL.n_stage: "N-category",
}
short_col_map = {tpl[-1]: val for tpl, val in col_map.items()}

raw = utils.load_patient_data(paths.data)
subdata = raw[col_map.keys()]
subdata.columns = subdata.columns.get_level_values(2)
subdata = (
  subdata
  .rename(columns=short_col_map)
  .reset_index(drop=True)
)

def n0_mean(n_stage: pd.Series) -> float:
  """Compute portion of N0 patients."""
  return (n_stage == 0).mean()

def early_mean(t_stage: pd.Series) -> float:
  """Compute early T-category portion."""
  return t_stage.isin([0,1,2]).mean()

grouped = (
  raw.groupby(
    by=COL.inst,
  ).aggregate(**{
    "Total": pd.NamedAgg(column=COL.age, aggfunc="count"),
    "Age (median)": pd.NamedAgg(column=COL.age, aggfunc="median"),
    "Neck Dissection": pd.NamedAgg(column=COL.nd, aggfunc="mean"),
    "N0": pd.NamedAgg(column=COL.n_stage, aggfunc=n0_mean),
    "Early T-Cat.": pd.NamedAgg(column=COL.t_stage, aggfunc=early_mean),
    "Mid. Ext.": pd.NamedAgg(column=COL.midext, aggfunc="mean"),
  }).convert_dtypes()
)
grouped.index.name = "Institution"
(
  grouped
  .reset_index()
  .style
  .format(
    formatter="{:>.0%}",
    subset=["Neck Dissection", "N0", "Early T-Cat.", "Mid. Ext."],
  )
  .apply(
    func=right_align,
    axis="index",
    subset=grouped.columns[0:],
  )
  .hide()
)
```

## Consensus on Involvement Status {#sec-data-consensus}

Pathological involvement is thus available only for a subset of the patients and only for the dissected levels. For the remaining patients, only clinical involvement is available. For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. These values are tabulated in @tbl-spec-sens.

This consensus was formed as follows: Suppose the involvement status $X_v$ of LNL $v$ was assessed using different diagnostic modalities $\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}$, each characterized by their own pair of sensitivity and specificity values $s_N^{\mathcal{o}}$ and $s_P^{\mathcal{o}}$, with $\mathcal{o} \in \mathcal{O}$. Then we have $|\mathcal{O}|$ observations $z_v^{\mathcal{o}} \in \left[ 0, 1 \right]$, where 0 stands for "healthy" and 1 for "involved". We can then compute the most likely true involvement $X_v$ using the likelihood function

$$
\begin{aligned}
\ell \left( X_v \mid \{ z_v^{\mathcal{o}} \}_{\mathcal{o} \in \mathcal{O}} \right) = \prod_{\mathcal{o} \in \mathcal{O}}
\left( 1 - X_v \right) \cdot &\left[ z_v^{\mathcal{o}} \cdot \left( 1 - s_P^{\mathcal{o}} \right) + \left( 1 - z_v^{\mathcal{o}} \right) \cdot s_P^{\mathcal{o}} \right] \\
+ X_v \cdot &\left[ z_v^{\mathcal{o}} \cdot s_N^{\mathcal{o}} + \left( 1 - z_v^{\mathcal{o}} \right) \cdot (1 - s_N^{\mathcal{o}}) \right]
\end{aligned}
$$

We now assume the true state $X_v$ to take on the value 1 if $\ell \left( X_v = 1 \mid \ldots \right) > \ell \left( X_v = 0 \mid \ldots \right)$ and 0 otherwise. For example, if we have $z_\text{II}^\text{CT} = 0$ and $z_\text{II}^\text{MRI} = 1$ we would compute the following likelihoods:

$$
\begin{aligned}
\ell \left( X_\text{II} = 1 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= \left( 1 - s_N^\text{CT} \right) \cdot s_N^\text{MRI} = 15.39\% \\
\ell \left( X_\text{II} = 0 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= s_P^\text{CT} \cdot \left(1 - s_N^\text{MRI}\right) = 14.44\%
\end{aligned}
$$

In this example, we would thus assume the true state to be involved ($X_\text{II} = 1$).

This method of computing a consensus also ensures that the pathology reports always override any conflicting clinical diagnosis, due to pathology's high sensitivity and specificity.

| Modality  | Specificity | Sensitivity |
| :-------- | ----------: | ----------: |
| CT        |         76% |         81% |
| PET       |         86% |         79% |
| MRI       |         63% |         81% |
| FNA       |         98% |         80% |
| pathology |        100% |        100% |

: Specificity and sensitivity values from the literature [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. {#tbl-spec-sens}

## Data Availability

The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023. The Data-in-Brief publications contain a detailed description of the datasets and the data format. However, these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.

## Patterns of Contralateral Involvement {#sec-data-strat}

These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and involvement. In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by risk factors such as T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal line. We also show in the fourth subfigre that T-category and ipsilateral involvement are not just rpedictive of contralateral involvement because they correlate with midline extension. A similar but more complete stratification is also tabulated in the appendix in @tbl-data-strat.

![Contralateral involvement stratified by T-category (top left panel), the number of metastatic LNLs ipsilaterally (top right panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (bottom left panel). Additionally, in the bottom right, we show that even for lateralized tumors, T-category and ipsilateral involvement are correlated with contralateral metastasis by comparing the contralateral involvement prevalence for selected scenarios that vary in their T-category and ipsilateral involvement extent.](figures/fig_data_strat.svg){#fig-data-strat}

The top left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement). This is simply because T-category may on average be considered a surrogate for the time between onset of disease and diagnosis. I.e., a patient with a T4 tumor was -- on average -- diagnosed later than a patient with a T1 tumor. Thus, the former did have more time to develop metastases.

Similarly, ipsilateral involvement correlates with contralateral metastasis. The tumor of a patients with many metastases in ipsilateral LNLs was probably able to spread for longer (or faster) compared to a tumor in a patient with no nodal disease. This, too, may therefore be considered a surrogate for the duration of the disease. In addition, it has been hypothesized that bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side. <!-- citation needed? -->

The bottom left panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors. This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline. Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion. Which in turn is only possible when the primary tumor is close enough to the mid-sagittal line or crosses it. <!-- citation(s) needed? or is this intuition good enough? -->

The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated. For example, midline extension was present in {{< var data.late_with_midext_percent >}}% of advanced T-category tumors but only in {{< var data.early_with_midext_percent >}}% of early T-category tumors. The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors. However, the bottom right panel of @fig-data-strat indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors. The figure considers only patients with lateralized tumors not extending over the midline. For early T-category and healthy ipsilateral levels I-V, only {{< var data.early_ipsin0_cII_percent >}}% ({{< var data.early_ipsin0_cII_match >}} out of {{< var data.early_ipsin0_cII_total >}} patients) shows involvement of contralateral level II. This increases to {{< var data.early_ipsiII_cII_percent >}}% ({{< var data.early_ipsiII_cII_match >}} out of {{< var data.early_ipsiII_cII_total >}} patients) if ipsilateral level II is involved, to {{< var data.early_ipsiIIandIII_cII_percent >}}% ({{< var data.early_ipsiIIandIII_cII_match >}} out of {{< var data.early_ipsiIIandIII_cII_total >}} patients) if ipsilateral levels II and III are involved, and to {{< var data.late_ipsiIIandIII_cII_percent >}}% ({{< var data.late_ipsiIIandIII_cII_match >}} out of {{< var data.late_ipsiIIandIII_cII_total >}} patients) for advanced T-category with ipsilateral levels II and III involved.

## Requirements for a Bilateral Model {#sec-requirements}

Based on the observations of the [previous section](#sec-data-strat), any model to predict the risk for contralateral nodal involvement, should be able to describe the following:

1. More advanced T-category should lead to higher risk for nodal disease. One approach to achieve this via the expected time of diagnosis has already been developed in the form of a hidden Markov model [@ludwig_hidden_2021].
2. The degree of ipsilateral involvement should give the model information on the time that may have passed between onset and diagnosis of the disease. This should come in addition to what can be inferred about the time from T-category alone.
3. A tumor that extends over the mid-sagittal line should yield contralateral metastases with much higher probability.

Over the course of this work, we will first briefly recap the HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only. Then, we intuitively extend it to include the contralateral side as well in @sec-ext-to-contra. In this section, we also introduce a way of modelling the tumor's midline extension as a random variable (@sec-midline) and lastly talk about how it may affect the contralateral spread in @sec-params-symmetry.

<!-- TODO: Hint at methods and results... -->

@@

This paper builds on the previously developed unilateral model for ipsilateral lymph node involvement presented in [cite 2024 graph extension paper]. In this section we provide a brief recap of the unilateral model to introduce the notation needed to extend the framework to a bilateral model describing both ipsilateral and contralateral lymph node involvement in @sec-ext-to-contra. For further details in the ipsilateral model, the reader is referred to the earlier publications [@ludwig_hidden_2021, cite 2024 graph extension paper]

<!--
Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019]. We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2021] to allow an intuitive inclusion of T-category into the predictions. We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.
-->

We model a patient's state of involvement at an abstract time-step $t$ as a vector of hidden binary random variables:

$$
\mathbf{X}[t] = \begin{pmatrix} X_v[t] \end{pmatrix} \qquad v \in \left\{ 1, 2, \ldots, V \right\}
$$ {#eq-state-def}

Here, $V$ is the number of LNLs the model considers. The values a LNL's hidden binary random variables may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some form of tumor presence (i.e., occult or clinically detected). Since the state vector $\mathbf{X}[t]$ is $V$-dimensional and binary, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$. In addition, each LNL is associated with an observed binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that suspicious lymph nodes were detected that are deemed metastatic.

Any hidden Markov model is fully described by three quantities:

1. A starting state $\mathbf{X}[t=0]$ at time $t=0$ just before the patient's tumor formed. In our case, this is always the state where all LNLs are still healthy $\boldsymbol{\xi}_0$.
2. The *transition matrix*
   $$
   \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
   $$ {#eq-trans-matrix}
   where the value at row $i$ and column $j$ represents the probability to transition from state $\boldsymbol{\xi}_i$ to $\boldsymbol{\xi}_j$ during the time-step from $t$ to $t+1$. Note that we prohibit self-healing, meaning that during a transition, no LNL may change their state from $X_v[t]=1$ to $X_v[t+1]=0$. Consequently, many elements of the transition matrix are zero.
3. Lastly, the *observation matrix*
   $$
   \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
   $$ {#eq-obs-matrix}
   where in row $i$ and at column $j$ we find the probability to *observe* a lymphatic involvement pattern $\mathbf{Z} = \boldsymbol{\zeta}_j$, given that the true (but hidden) state of involvement at the time of diagnosis $t_D$ is $\mathbf{X}[t_D] = \boldsymbol{\xi}_i$.

The transition matrix $\mathbf{A}$ is parametrized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network. Directed arcs from the primary tumor to a LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step. Directed arcs from a LNL $v$ to a LNL $r$ are associated with a probability $t_{vr}$ for the tumor to progress from one LNL to the next. In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [cite 2024 graph extension paper].

![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region. Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables. Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step. The orange arcs represent the sensitivity and specificity of the observational modality (e.g. CT, MRI, pathology, ...).](static/full-graph.svg){#fig-full-graph width=40%}

Using the introduced quantities, we can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$. For later use, we define at this point the matrices of the dsitributions over all hidden states, given all time-steps $\boldsymbol{\Lambda}$:

$$
\boldsymbol{\Lambda} = P \left( \mathbf{X} \mid \mathbf{t} \right) = \begin{pmatrix}
\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^0 \\
\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^1 \\
\vdots \\
\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^{t_\text{max}} \\
\end{pmatrix}
$$ {#eq-lambda-matrix}

Where the $k$-th row in this matrix corresponds to the probability distribution over hidden states after $t=k-1$ time-steps.

At the time of the diagnosis $t_D$, we multiply the result with the observation matrix $\mathbf{B}$. We may then look up the likelihood of a patient presenting with the diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_i$ in the $i$-th entry of the final result. However, the remaining issue is that the value of $t_D$ is unknown, i.e. over how many time-steps the HMM should be evolved. We solve this problem by marginalizing over the time of diagnosis. Different distributions over the diagnosis times can then be choosen based on T-category. For instance, the mean of the time-prior to marginalize over the diagnosis time for early T-category patients $P\left( t_D \mid \text{early} \right)$ may be shifted towards earlier times than the one for advanced T-category patients $P\left( t_D \mid \text{early} \right)$. This gives us for example

$$
P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
$$

In this work, we use binomial distributions $\mathfrak{B} \left( t_D, p_{\text{T}x} \right)$ as time-priors which have one free parameter $p_{\text{T}x}$ for each group of patients we differentiate based on T-category. Also, we fix $t_\text{max} = 10$, which means that the expected number of time-steps from the onset of a patient's disease to their diagnosis is $\mathbb{E}\left[ t_D \right] = 10 \cdot p_{\text{T}x}$.

## Likelihood Function of the Unilateral Model

With the formalism introduced above, we can write the likelihood function for a patient to present with a diagnosis consisting of an observed state and a T-category $d = \left( \boldsymbol{\zeta}_i, \text{T}x \right)$ as follows:

$$
\ell = P \left( \mathbf{Z} = \boldsymbol{\zeta}_i \mid \text{T}x \right) = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A}^t \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
$$ {#eq-single-patient-llh}

Above, the quantity inside $\left[ \ldots \right]_i$ denotes the $i$-th component of the vector that is the result of the vector and matrix multiplications in the square brackets. Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.

The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-full-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors. In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories. Therefore, the parameter space of the unilateral model is:

$$
\boldsymbol{\theta} = \left( \left\{ b_v \right\}, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
$$ {#eq-param-space}

And it is our goal to infer the values of these parameters for a given dataset $\mathcal{D} = \left( d_1, d_2, \ldots, d_N \right)$ of OPSCC patients. The likelihood of these $N$ diagnoses is simply the product of their individual likelihoods as defined in @eq-single-patient-llh. For numerical reasons, we typically compute the data likelihood in log space:

$$
\log \mathcal{L} \left( \mathcal{D} \mid \boldsymbol{\theta} \right) = \sum_{i=1}^N \log \ell_i
$$ {#eq-log-likelihood}

The methodology we use to infer the model's parameters is detailed in @sec-sampling.

# Extension to a Bilateral Model {#sec-ext-to-contra}

A naive approach to model the contralateral lymphatic spread would be to simply employ two independent unilateral models as introduced in @sec-unilateral. During training, one could enforce that some parameters are shared between these two models, e.g. the parameterization of the distributions over diagnose times or the spread among the LNLs ($t_{vr}). However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement discussed in section {#sec-data-strat}. Two independent unilateral models would not describe the observation that contralateral involvement becomes more likely with more severe ipsilateral involvement.

<!-- This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement. Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category. But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III. -->

<!-- Additionally, we could think of an approach to incorporate the primary tumor's mid-sagittal extension as a risk factor. -->

Thus, we extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously over time. To achieve that, we start by writing down the posterior distribution of involvement an analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:

$$
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)}
$$ {#eq-bilateral-bayes}

For the sake of brevity, we omit the dependency on the parameters and the T-category here.

The probability of the diagnoses given a hidden state factorises: $P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)$, and the two factors are described through observation matrices $\mathbf{B}^\text{i}$ and $\mathbf{B}^\text{c}$.

The term representing the model's prior probability of hidden involvement does not factorize. However, we assume that there is no direct lymph drainage from an ipsilateral LNL to a contralateral LNL. No major lymph vessels cross the mid-sagittal plane. In the graphical model, this means that we assume no directed arcs between ipsilateral and contralateral LNLs, i.e. tumor spread to the contralateral side is due to the primary tumor alone. We can thus write the joint probability $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)$ as a factorising sum:

$$
\begin{aligned}
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) &= \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid t \right) \\
&= \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
\end{aligned}
$$ {#eq-bilateral-marginal}

This assumption makes intuitive sense: The state of the ipsilateral and contralateral sides of the lymphatic network evolves independently over time as there are no major lymph vessels crossing the midline. However, both sides are coupled via time. Qualitatively speaking, according to the model in {#eq-bilateral-marginal}, a joint state with severe contralateral involvement and limited ipsilateral involvement would be unlikely, because severe contralateral involvement becomes likely only at later time steps when limited ipsilateral involvement becomes unlikely.

<!-- There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.-->

Using {#eq-bilateral-marginal} along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:

$$
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
$$ {#eq-bilateral-marginal-algebra}

## Parameter Symmetries {#sec-params-symmetry}

In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters, i.e., the ipsi- and contralateral spread rates are entirely different. However, using three sensible assumptions, we can reduce the parameter space by sharing some parameters between the two sides:

>>>>

---
title: "A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location"
abstract: |
  Purpose: to be done

  Methods: to be done

  Results: to be done

  Conclusions: to be done
---

# Introduction
Head and Neck Squamous Cell Carcinoma (HNSCC) are known to spread through the lymphatic system and form metastases in lymph nodes [@mukherji_cervical_2001, @shah_patterns_1990]. To avoid nodal recurrences, lymph node levels (LNL) at risk of harbouring occult metastases are electively irradiated. Guidelines for different tumor locations [@biau_selection_2019] are based on the overall prevalence of nodal disease as reported in the literature [@mukherji_cervical_2001, @shah_patterns_1990].
To personalize the prediction of the risk of occult metastases, given a patient's individual diagnosis, we first published a large, multi-centric dataset where the lymphatic involvement per LNL is available for each patient[@ludwig_dataset_2022, @ludwig_multi-centric_2023].
Subsequently we published an interpretable hidden Markov model (HMM), trained with the data to predict the risk for occult nodal disease, given an individual patient's diagnosis [@ludwig_hidden_2021].
A personalized risk prediction may allow clinicians to safely reduce the elective clinical target volume (CTV-N) and reduce side-effects that degrade the patient's quality of life without compromising treatment efficacy [@batth_practical_2014].

So far we trained different models for different tumor locations such as oropharynx and oral cavity tumors. However, this approach does not describe differences in lymphatic spread between different subsites within the oropharynx and oral cavity. This resulted in distorted predictions for each subsite, as the spread patterns were pooled into one single model .To address this issue, we present an approach using mixtures of HMMs. The intuition is that the lymphatic spread of a tumor that lies anatomically at the border of oropharynx and oral cavity (e.g. tumors in the palate) may be described by a mixture of different models.

@@

@@

# Mixture Model for Lymphatic Spread {#sec-mixture}

Since primary tumors at different subsites have different patterns of lymphatic spread patterns one would need to either generalize over several subsites as introduced in [@ludwig_dynamic_2021], where the HMM has been introduced for oropharyngeal cancer, or train a distinct model for each subsite. The former approach would lead to less precise predictions for each subsite, especially those subsites with few patients. Training one HMM per subsite would be rather time intensive and lead to large uncertainties for subsites with only few patients. Further we would not exploit the anatomical similarities which are expected for anatomically nearby locations.

Therefore we introduce a mixture model which combines all subsites in a singular model exploiting anatomical proximities.
